---
title: "Munkres Assignment Algorithm"
author: "Wolfgang Scherrer"
date: "23 July 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#bibliography: '`r system.file("REFERENCES.bib", package="rationalmatrices")`'
```


Given an $m\times n$ "cost" matrix $C$, find permutations of rows and columns such that (after the permutation) 

$$
\sum_{i=}^{\min(m,n)} C_{ii}
$$
is minimal. The cost matrix is often interpreted as follows. Consider $m$ agents and $n$ jobs and let $C_{ij}$ be the cost for assigning job $j$ to agent $i$. The task is to find an assignment of agents to jobs with 
the minimal total cost. Each agent can perform only one job, each job can be assigned only to one agent and 
we have to assign $\min(m,n)$ jobs. 

*Munkres'* Assignment Algorithm (sometimes also called *Hungarian Algorithm*) solves this 
problem in polynomial time. 

References: [@Munkres1956] ([@BourgeoisLassalle1971] extension for the non-square case) and 
http://csclab.murraystate.edu/~bob.pilgrim/445/munkres.html.


* In the package this algorithm is just used for *matching* poles and zeroes.

```{r}
# this is a noisy (re)implementation for debugging
munkres_debug = function(C, silent = FALSE) {
  # define some helper functions.
  # these helper functions are defined "inside" of munkres()
  # such that they have access to the environment of munkres().
  
  # print the current state and eventually what to next.
  print_state = function(next_step = NULL) {
    if (!silent) {
      # browser()
      rownames(state$C) = state$row_is_covered
      colnames(state$C) = state$col_is_covered
      state$C[state$stars] = NA_real_
      state$C[state$primes] = Inf
      print(state$C)
      if (!is.null(next_step)) {
        cat('next step:', next_step, '\n')
      }
    }
  } 
  
  # mark zeroes as "stars"
  # each row, each column may contain at most one starred zeroe. 
  # if m (<= n) zeroes are starred, then we have found the optimal assignment
  star_zeroes = function() {
    # browser()
    # C is square or wide: (m-by-n) and m <= n.
    m = nrow(state$C)
    state$stars = matrix(0, nrow = 0, ncol = 2)
    cc = logical(ncol(state$C)) # cc[j] is set to TRUE if the column j contains a star
    for (i in (1:m)) {
      j = which((state$C[i, ] == 0) & (!cc))
      if ( length(j) > 0) {
        cc[j[1]] = TRUE
        state$stars = rbind(state$stars, c(i,j[1]))
      }
    }
    
    state ->> state
    next_step = 'cover cols with stars'
    print_state(next_step)
    
    return(next_step)
  }
  
  # cover/mark the columns which contain a starred zeroe, 
  # if there are m (<= n) starred zeroes, then these 
  # starred zeroes describe the optimal assignment 
  cover_cols_with_stars = function() {
    # browser()
    m = nrow(state$C)
    state$col_is_covered[] = FALSE
    state$col_is_covered[state$stars[, 2]] = TRUE
    
    state ->> state
    if (sum(state$col_is_covered) == m) {
      next_step = 'done'
    } else {
      next_step = 'prime zeroes'
    }
    print_state(next_step)
    
    return(next_step)
  }
   
  # find a zeroe within the non-covered entries of C
  find_zeroe = function() {
    state$C[state$row_is_covered, ] = 1 # in order to exclude the covered rows from the search
    state$C[, state$col_is_covered] = 1 # in order to exclude the covered cols from the search
    # browser()
    if (any(state$C == 0)) {
      i = which.max(apply(state$C == 0, MARGIN = 1, FUN = any))
      j = which.max(state$C[i, ] == 0)
      return(c(i,j))
    } else {
      return(NULL)
    }
  }
  

  # find a zeroe within the non-covered entries of C and mark it as "primed". 
  # primed zeroes are "candidates" to replace starred zeroes 
  prime_zeroes = function() {
    # iter = 1
    # while ((TRUE) && (iter<=100)) {
    while (TRUE) {
      # the function "returns", if no zeroe can be found, or if there is no 
      # star in the respective row. otherwise the row is "covered".
      # hence the while loop will stop after at most m iterations. 
      ij = find_zeroe()
      if (is.null(ij)) {
        state ->> state
        next_step = 'augment path'
        print_state(next_step)
    
        return(next_step)
      } else {
        if (!silent) cat('prime zeroe', ij, '\n')
        state$primes = rbind(state$primes, ij)
        i = ij[1]
        k = which(state$stars[, 1] == i) # find column with star in the i-th row 
        if ( (length(k) == 0)) {
          state$zigzag_path = matrix(ij, nrow = 1, ncol = 2)
          state ->> state
          next_step = 'make zigzag path'
          print_state(next_step)
    
          return(next_step)
        } else {
          j = state$stars[k[1], 2]
          state$row_is_covered[i] = TRUE  # cover the i-th row 
          state$col_is_covered[j] = FALSE # uncover the j-th column
          state ->> state
        }
      }
      # iter = iter + 1
    }
    # if (iter >= 100)  stop('did not converge')
  }
  
  # create a path, which connects alternating primed and starred zeroes. 
  # 
  # the path starts and ends with a prime zeroe. 
  # at the end, the starred zeroes of the path are "unstarred" and the primed zeroes 
  # get starred zeroes. So we construct an additional starred zeroe!
  #
  make_zigzag_path = function() {
    pos = 1 # current length of zigzag path 
    done = FALSE 
    # iter = 1
    # while ((!done) && (iter <= 100)) {
    while (!done) {
      # the while loop stops after at most m iterations, since ...
      
      # the current zeroe is a primed zero 
      # find a starred zero in this column
      k = which(state$stars[, 2] == state$zigzag_path[pos, 2]) 
      if (length(k) == 0) {
        done = TRUE   # if we cannot find a starred zeroe, the zigzag path is finished
      } else {
        pos = pos + 1 # add this starred zeroe to the path
        state$zigzag_path = rbind(state$zigzag_path, c(state$stars[k,]))
      }
      
      if (!done) {
        # find a primed zero in this row 
        k = which(state$primes[, 1] == state$zigzag_path[pos, 1]) 
        if (length(k) == 0) {
          stop('this should not happen (1)')
        } else {
          pos = pos + 1 # add this primed zeroe to the zigzag path
          state$zigzag_path = rbind(state$zigzag_path, c(state$primes[k,]))
        }
      }
      # iter = iter + 1
    }
    # if (iter >= 100) stop('did not converge')
    if (!silent) print(state$zigzag_path)

    # modify stars/primes along the path 
    for (i in (1:pos)) {
      if ((i %% 2) == 1) {
        # primed zeroe -> starred zeroe 
        state$stars = rbind(state$stars, state$zigzag_path[i,])
      } else {
        # starred zeroes get unstarred 
        k = which( (state$stars[, 1] == state$zigzag_path[i, 1]) & (state$stars[, 2] == state$zigzag_path[i, 2]) )
        if (length(k == 1)) {
          state$stars = state$stars[-k, , drop = FALSE]
        } else {
          cat('i=', i, '\n')
          print(state$zigzag_path)
          print(state$stars)
          print(state$primes) 
          print(k)
          stop('this should not happen (2)')
        }
      }
    }
    state$row_is_covered[] = FALSE   # uncover all rows
    state$col_is_covered[] = FALSE   # uncover all columns 
    state$primes = matrix(0, nrow = 0, ncol = 2) # delete all primed zeroes
    
    state ->> state
    next_step = 'cover cols with stars'
    print_state(next_step)
    
    return(next_step)
  }

  # augment path 
  # this step is executed, if the matrix C does not contain "uncovered" zeroes. 
  # let m be the minimum of the non-covered elements. 
  # substract m from the uncovered elements and add m to the elements which 
  # are double covered (column and ro is covered).
  augment_path = function() {
    m = min(state$C[!state$row_is_covered, !state$col_is_covered])
    state$C[!state$row_is_covered, !state$col_is_covered] = 
      state$C[!state$row_is_covered, !state$col_is_covered] - m
    state$C[state$row_is_covered, state$col_is_covered] = 
      state$C[state$row_is_covered, state$col_is_covered] + m

    state ->> state
    next_step = 'prime zeroes'
    print_state(next_step)
    
    return(next_step)
  }

  C0 = C
  m = nrow(C)
  n = ncol(C)

  # convert to "wide" matrix
  if (m > n) {
    transposed = TRUE
    C = t(C)
    m = nrow(C)
    n = ncol(C)
  } else {
    transposed = FALSE
  }

  # adding/substracting a scalar to a row or column of C does not
  # change the optimal assignment(s). 
  # (However, the minimal total cost is changed.)
  # 
  # substract the row minima (from the respective rows) and the 
  # column minima (from the respective columns). 
  # this gives a matrix with nonnegative entries and at least 
  # one zero in each column and each row. 
  C = C - matrix(apply(C, MARGIN = 1, FUN = min), m, n)
  # C = C - matrix(apply(C, MARGIN = 2, FUN = min), m, n, byrow = TRUE)
  
  # construct a "state" variable, which stores the current state. 
  state = list(C = C,
               row_is_covered = logical(m),
               col_is_covered = logical(n),
               stars = matrix(0, nrow = 0, ncol = 2),
               primes = matrix(0, nrow = 0, ncol = 2),
               zigzag_path = matrix(0, nrow = 0, ncol = 2))
  
  next_step = star_zeroes()

  # iter = 1
  done = FALSE
  while (!done) {
    # print(iter)
    
    if (next_step == 'cover cols with stars') {
      next_step = cover_cols_with_stars()
      if (next_step == 'done') break   # we are done!!!
    }
    
    if (next_step == 'prime zeroes') {
      next_step = prime_zeroes()
    }
    
    if (next_step == 'make zigzag path') {
      next_step = make_zigzag_path()
    } 
    if (next_step == 'augment path') {
      next_step = augment_path()
    }
    # iter = iter + 1
    
  }

  # browser()  
  state$C = C0
  
  # starred zeroes represent the optimal matching 
  state$match = state$stars
  if (transposed) state$match = state$match[, c(2,1)]
  state$match = state$match[order(state$match[, 1]), ]
  state$cost = sum(state$C[state$match])
  return(state[c('match','cost','C')])
} 
```

```{r}
C = matrix(c (0, 0, 0, 0, 0, 1, 3, 3, 0, 5, 5, 9, 0, 1, 3, 7), nrow = 4, ncol = 4, byrow = TRUE) + 10
out = munkres_debug(C)
```


```{r}
C = matrix(c (2, 3, 3, 3, 2, 3, 3, 3, 2), nrow = 3, ncol = 3, byrow = TRUE)
out = munkres_debug(C)
```


```{r}
C = matrix(c (82, 83, 69, 92, 77, 37, 49, 92, 11, 69, 5, 86, 8, 9, 98, 23), nrow = 4, ncol = 4, byrow = TRUE)
out = munkres_debug(C)
```

```{r}
C = matrix(c (90, 75, 75, 80, 35, 85, 55, 65, 125,95, 90, 105, 45, 110, 95, 115), nrow = 4, ncol = 4, byrow = TRUE)
out = munkres_debug(C)
```


```{r}
C = matrix(c (2500, 4000, 3500, 4000, 6000, 3500, 2000, 4000, 2500), nrow = 3, ncol = 3, byrow = TRUE)
out = munkres_debug(C)
```

```{r}
m = sample(3:5, 1)
n = sample(3:5, 1) 
cat(m, n, '\n')

C = matrix(sample(1:10, m*n, replace = TRUE), nrow = m, ncol = n)
out = munkres_debug(C)
print(C)
print(out$match)
print(out$cost)

if (m != n) {
  if (m < n) C2 = rbind(C, matrix(max(C), nrow = n-m, ncol = n))
  if (m > n) C2 = cbind(C, matrix(max(C), nrow = m, ncol = m-n))
  out2 = munkres_debug(C2, TRUE)
  match = out2$match 
  match = match[(match[,1] <= m) & match[,2] <=n, ]
  print(match)
  print(sum(C[match]))
}
```
  

