---
title: "Test Multiplication of Matrix Polynomials"
author: "WS"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rationalmatrices)
```

Vergleich die "alte" Implementation der Multiplikation von zwei Matrix Polynomen 
(polm, bzw. pseries Objekten) mit einer neuen Implementation.

* Die neu Implementation vermeidet die "Doppelschleife" und ist daher etwas schneller 
für Polynome/pseries großer Ordnung. 


# "old" implementations 


```{r}
# Internal functions acting on ARRAYS for matrix multiplication ####
# multiplication of matrix polynomials
# this function performs only basic checks on the inputs!
mmult_poly = function(a, b) {
  # a,b must be two compatible arrays
  da = dim(a)
  db = dim(b)
  if (da[2] != db[1]) stop('arguments are not compatible')
  # if any of the arguments is an empty polynomial, or a polynomial of degree -1
  if (min(c(da,db)) == 0) return(array(0, dim = c(da[1], db[2], 0)))
  
  # skip zero leading coefficients
  if (da[3] > 0) {
    a = a[ , , rev(cumprod(rev(apply(a == 0, MARGIN = 3, FUN = all)))) == 0, drop = FALSE]
    da = dim(a)
  }
  # skip zero leading coefficients
  if (db[3] > 0) {
    b = b[ , , rev(cumprod(rev(apply(b == 0, MARGIN = 3, FUN = all)))) == 0, drop = FALSE]
    db = dim(b)
  }
  # if any of the arguments is an empty polynomial, or a polynomial of degree -1
  if (min(c(da,db)) == 0) return(array(0, dim = c(da[1], db[2], 0)))
  
  pa = da[3] - 1
  pb = db[3] - 1
  
  x = array(0, dim = c(da[1], db[2], pa + pb + 1))
  # the 'convolution' of the coefficients is computed via a double loop
  # of course this could be implemented more efficiently!
  for (i in (0:(pa+pb))) {
    for (k in iseq(max(0, i - pb), min(pa, i))) {
      x[,,i+1] = x[,,i+1] + 
        matrix(a[,,k+1], nrow = da[1], ncol = da[2]) %*% 
          matrix(b[,,i-k+1], nrow = db[1], ncol = db[2])
    }
  }
  return(x)
}

# internal function
# multiplication of two impulse response functions
# this function performs only basic checks on the inputs!
# almost equal to mmult_poly
mmult_pseries = function(a, b) {
  # a,b must be two compatible arrays
  da = dim(a)
  db = dim(b)
  if (da[2] != db[1]) stop('arguments are not compatible')
  # if any of the arguments is an empty pseries, or a pseries with lag.max = -1
  if (min(c(da,db)) == 0) return(array(0, dim = c(da[1], db[2], min(da[3], db[3]))))
  
  lag.max = min(da[3], db[3]) - 1
  # truncate to the minimum lag.max
  # a = a[ , , 1:(lag.max+1), drop = FALSE]
  # b = b[ , , 1:(lag.max+1), drop = FALSE]
  
  x = array(0, dim = c(da[1], db[2], lag.max + 1))
  # the 'convolution' of the impulse response coefficients is computed via a double loop
  # of course this could be implemented more efficiently!
  for (i in (0:lag.max)) {
    for (k in (0:i)) {
      x[,,i+1] = x[,,i+1] + 
        matrix(a[,,k+1], nrow = da[1], ncol = da[2]) %*% 
          matrix(b[,,i-k+1], nrow = db[1], ncol = db[2])
    }
  }
  return(x)
}
```


# "new" implementation

* avoid double loop
* however, in order to simplify the code, we do not 
  care about trailing/leading zero coefficients 

```{r}
convolve_3D = function(a, b, truncate = FALSE) {
  
  # a,b must be two compatible arrays
  # a is an (m,n) dimensional polynomial of degree p
  d = dim(a)
  m = d[1]
  n = d[2]
  p = d[3] - 1
  
  # d is an (n,o) dimensional polynomial of degree q
  d = dim(b)
  if (d[1] != n) stop('arguments are not compatible')
  o = d[2]
  q = d[3] - 1
  
  # output c = a*b is an (m,o) dimensional polynomial of degree r
  
  
  if (truncate) {
    # multiplication of two "pseries" objects, 
    
    # lag.max = minimum of the number of lags of a,b
    r = min(p, q)
    
    # if any of the arguments is an empty pseries, or a pseries with lag.max -1
    # ensure that lag.max = r >= 0
    if (min(c(m, n, o, r+1)) == 0) return(array(0, dim = c(m, o, max(r, 0)+1)))

    if (p > r) a = a[,,1:(r+1), drop = FALSE]
    if (q > r) b = b[,,1:(r+1), drop = FALSE]
    p = r
    q = r
  } else {
    # multiplication of two polynomials
    
    # if any of the arguments is an empty polynomial, or a polynomial of degree -1
    if (min(c(m, n, o, p+1, q+1)) == 0) return(array(0, dim = c(m, o, 0)))
    
    # degree = sum of the two degrees 
    r = p + q
  }
  
# cat('truncate', truncate,'\n')  
  if (p <= q) {
    c = matrix(0, nrow = m, ncol = o*(r+1))
    b = matrix(b, nrow = n, ncol = o*(q+1))
    for (i in (0:p)) {
      j1 = i
      j2 = min(j1+q+1, r+1)
# cat('a*b', i, j1, j2, '|', dim(c), ':', (j1*o + 1), (j2*o), 
#      '|', dim(b), ':', 1, ((j2-j1)*o), '\n')
      c[ , (j1*o + 1):(j2*o)] = 
        c[ , (j1*o + 1):(j2*o)] + 
          matrix(a[,,i+1], nrow = m, ncol = n) %*% b[,1:((j2-j1)*o) , drop = FALSE]
    }
    c = array(c, dim = c(m,o,r+1))
    return(c)
  } else {
    # first compute b' * a'
    c = matrix(0, nrow = o, ncol = m*(r+1))
    a = matrix(aperm(a, c(2,1,3)), nrow = n, ncol = m*(p+1))
    b = aperm(b, c(2,1,3)) 
    for (i in (0:q)) {
      j1 = i
      j2 = min(j1+p+1, r+1)
# cat('b*a', i, j1, j2, '|', dim(c), ':', (j1*m + 1), (j2*m), 
# '|', dim(a), ':', 1, ((j2-j1)*m), '\n')
      c[ , (j1*m + 1):(j2*m)] = 
        c[ , (j1*m + 1):(j2*m)] + 
          matrix(b[,,i+1], nrow = o, ncol = n) %*% a[,1:((j2-j1)*m) , drop = FALSE]
    }
    c = array(c, dim = c(o,m,r+1))
    c = aperm(c, c(2,1,3))
  }
  return(c)
}
```


# Test 

```{r}
n.sim = 1000
comp.time0 = 0
comp.time1 = 0
for (i in (1:n.sim)) {
  m = sample(0:5,1)
  n = sample(0:5,1)
  o = sample(0:5,1)
  p = sample(0:50,1)
  q = sample(0:50,1)
  
  a = array(rnorm(m*n*(p+1)), dim = c(m,n,p+1))
  b = array(rnorm(n*o*(q+1)), dim = c(n,o,q+1))
  
  start = Sys.time()
  # polynomial multiplication
  c0a = mmult_poly(a,b)
  c0b = mmult_pseries(a,b)
  comp.time0 = comp.time0 + (Sys.time() - start)
  
  start = Sys.time()
  c1a = convolve_3D(a,b)
  c1b = convolve_3D(a,b, TRUE)
  comp.time1 = comp.time1 + (Sys.time() -start)

  testthat::expect_equal(c0a,c1a)
  testthat::expect_equal(c0b,c1b)
}


print(comp.time0)
print(comp.time1)
```

